Approximation as Differentiation

### 2025-05-24 main/introduction.tex
siskind08.

Provenance tracking aims to answer questions like ``Where did these data come from?'', ``Why are these data in the output?'' and ``How were these data computed?''. Provenance tracking has a wide range of applications, from debugging and program comprehension to improving reproducibility and transparency in scientific workflows. In order to audit any computational process, we need robust and well-founded notions of provenance to track how data are used, how data are computed, and what data influenced a particular outcome. \emph{Program slicing}, first proposed by~\citet{weiser81}, is a collection of techniques for provenance tracking that attempts to take a run of a program and areas of interest in the output, and turn them into the subset of the input and the program that were responsible for generating those specific outputs.

@@

Previous work on \GPS uses an operational semantics of programs that generates a trace of each execution which can be interpreted after the fact to compute the Galois connection described above.

@@

\todo{explain why we want a more abstract view on \GPS}
There is a close analogy between \GPS and automatic differentiation for differentiable programs. We have already hinted at this in the description above, but let us now make it explicit.

>>>>

Data provenance aims to answer questions like ``Where did these data come from?'', ``Why are these data in the output?'' and ``How were these data computed?''. Provenance tracking has a wide range of applications, from debugging and program comprehension to improving reproducibility and transparency in scientific workflows. In order to audit any computational process, we need robust and well-founded notions of provenance to track how data are used, how data are computed, and what data influenced a particular outcome. \emph{Program slicing}, first proposed by~\citet{weiser81}, is a collection of techniques for provenance tracking that attempts to take a run of a program and areas of interest in the output, and turn them into the subset of the input and the program that were responsible for generating those specific outputs.

@@

@@

Previous work on \GPS uses an operational semantics of programs that generates a trace of each execution which can be rewound or replayed after the fact, to compute the Galois connection described above. It would be satisfying to have a denotational account of \GPS as well, for which traces would then offer a plausible implementation technique (perhaps one of several). As it happens there is a close analogy between \GPS and \emph{automatic differentiation} for differentiable programs~\cite{siskind08}, which points to a way to develop such an approach. We have already hinted at this in the description above, but let us now make it explicit.
